Variational Monte Carlo on bosonic systems

### 2018-02-27 project_1/article/article.tex
Add multi-dimensional importance sampling

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm (which will be discussed in due
        time). This algorithm is dependent on the drift force of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a transition
        probability and the acceptance probability.  The strength of the
        algorithm is that the transition algorithm need not be known. For
        example, the simplest case is to accept the new state, i.e., the new
        position for the random walker, if the ratio
        \begin{align}
            w &= \frac{\Psi_T(\vf{r}_{i + 1})}{\Psi_T(\vf{r}_{i})},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1)$.

\subsubsection{Importance Sampling}
A problem with the naïve Metropolis-Hastings sampling approach is that the sampling of the position space is done with
no regard for where we are likely to find a particle. This problem can be remedied through a \emph{importance sampling}.
It is reasonable to assume that the particles we erratically scatter in space are prone to movement towards the peaks
of the probability density as dictated by the wave function. Consider therefore the Fokker-Planck equation,
\begin{equation}
    \frac{\partial \psi}{\partial t} = D \frac{\partial }{\partial x}\left(\frac{\partial }{\partial x} - F \right) \psi,
    \label{eq:fokker_planck}
\end{equation}
which described the evolution in time of a probability density function, here exchanged for the wavefunction $\psi$.
Originaly an equation that models diffusion, we have a diffusion term $D$ and a drift force,
\begin{equation}
    F = \frac{2}{\psi} \frac{\partial \psi}{\partial x}.
\end{equation}

The one-particle model corresponding to the Fokker-Planck equation is the Langevin equation,
\begin{equation}
    \frac{\partial x}{\partial t} = D F(x) + \eta,
    \label{eq:langevin}
\end{equation}
where $\eta$ is a uniformly distributed stochastic variable. Solving Langevin's equation by Eurler's method gives a recursive relation for
the subsequent new positions of a particle,
\begin{equation}
    x_{t+\Delta t} = x_t + D F(x) \Delta t +\xi \sqrt{\Delta t},
\end{equation}
given a time step $\Delta t$\footnote{Bear in mind that \autoref{eq:langevin} is only valid as $\Delta t \to 0$, a property stemming from the use of Euler's method.} and a normally distributed stochastic variable $\xi$.

Now we need to change the acceptance probability of the metropolis algorithm to something that takes
the new sampling method into account,
\begin{equation}
    q(x_{\text{new}}, x) = \frac{G(x, x_{\text{new}}, \Delta t)\abs{\psi_T(x_{\text{new}})}^2}{G(x_{\text{new}}, x, \Delta t)\abs{\psi_T(x)}^2}
\end{equation}
where $G$ is the Green's function to the Fokker-Planck equation,
\begin{align}
    G&(x_{\text{new}}, x, \Delta x) \nonumber \\
    &= \frac{1}{(4\pi D \Delta t)^{3N/2}}\exp\left(-\frac{(x_{\text{new}} - x - D\Delta t F(x))^2}{4D\Delta t} \right).
\end{align}

>>>>

    \subsection{The drift force}
        A disadvantage in the use of the brute-force Metropolis algorithm is
        that we might be spending much computational resources in an
        uninteresting part of configuration space. To make smarter moves we will
        use the Metropolis-Hastings algorithm (which will be discussed in due
        time). This algorithm is dependent on the drift force of the system.
        \begin{align}
            \vf{F}(\vf{r})
            &=
            \sum_{k = 1}^N
            \vf{F}_k(\vf{r})
            =
            \sum_{k = 1}^N
            \frac{2\nabla_k\Psi_T(\vf{r})}{\Psi_T(\vf{r})}.
            \label{eq:drift_force}
        \end{align}
        Using this expression we are able to move towards parts of
        configuration space where the gradient increases or decreases yielding a
        better choice of movements. We will mainly be interested in the drift
        force of a single particle $k$.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a transition
        probability and the acceptance probability.  The strength of the
        algorithm is that the transition algorithm need not be known. For
        example, the simplest case is to accept the new state, i.e., the new
        position for the random walker, if the ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1)$.

        \subsubsection{Importance Sampling}
            A problem with the naïve Metropolis-Hastings sampling approach is
            that the sampling of position space is done with no regard for where
            we are likely to find a particle. This problem can be remedied
            through \emph{importance sampling}.  It is reasonable to assume that
            the particles we erratically scatter in space are prone to movement
            towards the peaks of the probability density as dictated by the wave
            function. Consider therefore the Fokker-Planck equation,
            \begin{align}
                \dpd[]{\Psi_T}{t}
                &=
                D \nabla\cdot
                \left(
                    \nabla
                    - \vf{F}
                \right) \Psi_T,
                \label{eq:fokker_planck}
            \end{align}
            which describes the evolution in time of a probability density
            function. In our case this is the trial wavefunction $\Psi_T$.
            Originaly an equation that models diffusion, we have a diffusion
            term $D$ and the drift force \autoref{eq:drift_force}. In our case
            the diffusion term $D$ is simply $1/2$ from the kinetic energy (in
            natural units).

            We use the Langevin equation to find the new position of the
            particle.
            \begin{align}
                \dpd[]{\vf{r}}{t}
                &=
                D\vf{F}(\vf{r}) + \vfg{\eta},
                \label{eq:langevin}
            \end{align}
            where $\vfg{\eta}$ is a uniformly distributed stochastic variable
            for each dimenion.  Solving Langevin's equation by Euler's method
            gives a recursive relation for the subsequent new positions of a
            particle.
            \begin{align}
                \vf{r}_{i + 1}
                &=
                \vf{r}_i + D\vf{F}(\vf{r}_i)\Delta t
                + \vfg{\xi}\sqrt{\Delta t},
            \end{align}
            for a given time step $\Delta t$
            \footnote{
                Bear in mind that \autoref{eq:langevin} is only valid as $\Delta
                t \to 0$, a property stemming from the use of Euler's method.
            }
            and a normally distributed stochastic variable $\vfg{\xi}$.

            Now we need to change the acceptance probability of the metropolis
            algorithm to something that takes the new sampling method into
            account.
            \begin{align}
                q(\vf{r}_{i + 1}, \vf{r}_i)
                &=
                \frac{
                    G(\vf{r}_{i + 1}, \vf{r}_{i}, \Delta t)
                    \left|\Psi_T(\vf{r}_{i + 1})\right|^2
                }
                {
                    G(\vf{r}_{i}, \vf{r}_{i + 1}, \Delta t)
                    \left|\Psi_T(\vf{r}_{i})\right|^2
                },
            \end{align}
            where $G$ is the Green's function of the Fokker-Planck equation
            given by
            \begin{align}
                G(\vf{r}_{i + 1}, \vf{r}_i, \Delta t)
                &=
                \exp\Biggl(
                    -\frac{\bigl[
                        \vf{r}_{i + 1}
                        - \vf{r}_{i}
                        - D\vf{F}(\vf{r}_i)\Delta t
                    \bigr]^2}{4D\Delta t}
                \Biggr)
                \nonumber \\
                &\qquad
                \times
                \frac{1}{(4\pi D\Delta t)^{dN/2}},
            \end{align}
            where $d$ is the dimensionality.
